Fractional Types

### 2016-12-18 typespp.tex
major revisions; sec 5 completely blank

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^/$}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^/$: Operational Semantics}

at end of previous section; add a new subsection explaining functor showing
equivalence of 1 and $\order{p} \times \iorder{p}$. Morita functor. Pullback condition. there is a
choice for eta but there is an extra condition that the choices are all
consistent. The big question is how to encode this global condition; side
effect; synchronization; etc. so naturally lead to pointed types and revise types

Only non-trivial pairs are created by eta; they create ``synchronized'' uses combinators $c^k$ and $1/c^k$. The rest of the combinators can only replace these combinators by $\isotwo$-equivalent ones. Assoc is subtle: what if we have $\langle \order{c_1} , \langle \iorder{c_2} , \order{c_2} \rangle \rangle$; naive assoc would allow us to create a pair with unrelated combinators which cannot be eliminated using epsilon. But if assoc is constrained to maintain the invariant that only ``synchronized'' uses of combinators are ever paired together, everything works great.

Idea is to change the types to be of the form $\Sigma_{\overline{k}}\mathbb{T}$
where we keep track of the values (pointed types). When we take a product we
concatenate the two sequences of integers.
\[
(\textrm{Fractional Types})  \mathbb{T} ::=
  \Sigma_k\order{c} \alt \Sigma_k\iorder{c} \alt
  \Sigma_{\overline{j},\overline{k}}\mathbb{T}_1 \circledast \mathbb{T}_2
\]

The operational semantics for all the primitive combinators is a
simple transliteration of Fig.~\ref{fig:opsem}. We omit the
straightforward implementation of prim and its inverse
below:

The interesting combinators operationally are the unit/counit combinators $\eta$
and $\epsilon$, and the synchronization combinators. Their semantics is defined
by the following pair of mutually recursive interpreters:

The unit combinators $\eta$ simply generate unit tangles, and since unit tangles
are equivalent to the \emph{identity}, they can be freely eliminated by the
counit combinators $\epsilon$. The two synchronization operations exchange the
tangled product with an iterate. The two interpreters satisfy several properties
described below: First they are congruences for the equivalence $≈$ on values.
We can further prove that both are \emph{reversible}.  Furthermore, we can show
two coherence conditions: first, that the reverse interpreter gives equivalent
results to the forward interpreter applied to a reverse combinator, and second,
that two equivalent combinators (as given by a 2-combinator) will evaluate to
equivalent values.

The operational definition of synchr⋆ might seem at odds with our claim that it is related to associativity while it looks like commutativity.  The way to understand it is as follows: if indeed all values were independent, it would be associativity.  However, in this case, $p // q$ is really two \emph{tangled} values, which we do not have access to.  To keep these synchronized and yet to achieve the given type, the only choice (operationally) is to swap.  This is where the ``action at a distance'' occurs.

The following is an example usage of zig-zag, which uses NOT combinator.

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{cc}
\begin{array}[t]{rlcl}
  \evalone{\unitetl/}{&\langle id^{k},\mathbb{V} \rangle} &=& \mathbb{V} \\
  \evalone{\unititl/}{& \mathbb{V}} &=& \langle id^{k_1},\mathbb{V} \rangle \\
  \evalone{\unitetr/}{&\langle \mathbb{V},id^{k} \rangle} &=& \mathbb{V} \\
  \evalone{\unititr/}{& \mathbb{V}} &=& \langle \mathbb{V},id^{k} \rangle\\
  \evalone{\swapt}{&\langle \mathbb{V}_1, \mathbb{V}_2 \rangle} &=& \langle \mathbb{V}_2, \mathbb{V}_1 \rangle \\
  \evalone{\assoclt/}{&\langle c^k , \langle 1/c^{k}, c^{k} \rangle\rangle} &=& \langle \langle c^k , 1/c^{k} \rangle, c^{k} \rangle \\
  \evalone{\assocrt/}{&\langle \langle c^k , 1/c^{k} \rangle, c^{k} \rangle} &=& \langle c^k , \langle 1/c^{k}, c^{k} \rangle\rangle \\
\end{array}
&
\begin{array}[t]{rlcl}
  \evalone{\eta-}{& id^{k_1}} &=& \langle 1/c^{k_2},c^{k_2} \rangle\\
  \evalone{\eta+}{& id^{k_1}} &=& \langle c^{k_2},1/c^{k_2} \rangle\\
  \evalone{\epsilon-}{& \langle 1/c^{k_1},c^{k_1} \rangle} &=& id^{k_2} \\
  \evalone{\epsilon+}{& \langle c^{k_1},1/c^{k_1} \rangle} &=& id^{k_2} \\
\end{array}

\end{array}\]
\caption{\label{fig:opsem-frac}$\Pi^/$ operational semantics}
}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Limitations}

Given any non-negative rational number, we can form a type whose cardinality is
that number. And yet, our types do not capture the full structure of the
non-negative rational numbers, as these form a commutative semifield. There are
a number of operations and properties which need to be added to complete this.
Most seem quite straightforward.  We can however single one out which may not
be: what we are missing is a multiplicative inverse for every type, and not just
$\order{p}$. In particular, we would like to form the type
$\iorder{(\iorder{p})}$ and have it be equivalent to $\order{p}$.  The
relationship between $\order{p}$ and $\iorder{p}$ in $\Pi^/$ reveals a pattern
that can be exploited to achieve this generalization. The multiplicative inverse
of $\order{p}$ is obtained by promoting the objects in $\order{p}$ to morphisms
in $1/\hash p$, which is a general process called \emph{delooping}. It is clear
we can go in the inverse direction by demoting morphisms to objects which is a
process called \emph{looping}. We leave a possible extension with a general
looping/delooping process to future work.

A second more immediate concern is to discover the abstract monadic and/or
comonadic structure underlying the operational semantics. It is clear that the
operational interpretation of fractional types involves a computational effect
that requires synchronization with the ``future''. Implementing this effect
using backtracking, reference cells, or dataflow constraints is possible but
each implementation choice is ad hoc. We conjecture that with the right monadic
and/or comonadic abstraction, the operational semantics will be simpler and
expose still richer properties.  We chose the dependently-typed dataflow
semantics because we are able to prove that it is reversible and coherent,
unlike our previous attempts with explicit effects, none of which were provably
reversible and/or coherent.

@@

We have presented a natural notion of \emph{fractional types} that
enriches a class of reversible programming languages in several
dimensions.  Further research might show how to use fractional
types in a conventional (i.e., irreversible) programming language, their full
potential is only achieved when the ambient language guarantees that
no information is created or erased.

The key semantic insight is that iterating a reversible program $p$ on
a finite type must eventually reach the identity in $\ord{p}$
steps. By being careful not to collapse proofs, each such reversible
program has $\ord{p}$ distinct proofs of reversibility and hence gives
rise to a groupoid with cardinality $\frac{1}{\ord{p}}$. Going from
this observation to a full programming language required several
difficult and subtle design choices which we have explored to produce
$\Pi^/$. The latter language has a natural denotational semantics
where types denote groupoids. Its operational semantics requires a
mechanism to express a computational effect which enables spatially
separated parts of the program to communicate in a way that is
reminiscent of entanglement in quantum mechanics. It is possible to
realize such an operational semantics using global reference cells,
backtracking, or other conventional technique. A more enlightening and
less ad hoc implementation encodes the required dependency in dataflow
constraints encoded in dependent types. The key idea is to generalize
the usual cartesian product to a \emph{tangled product} that allows the
components to interact at synchronization points.

Our fractional types extend the natural denotation of types from sets
to non-trivial groupoids but they only scratch the surface of the
tower of weak $\omega$-groupoids that is expressible in HoTT. A long
term goal of our research is to find natural type constructors
inspired by the rich combinatorial structure of weak
$\omega$-groupoids and that provide novel programming abstractions.

For an information-theoretic perspective on the language $\Pi$, we think of a
type containing $N$ values as an abstract system that has~$N$ distinguishable
states. According to the conventional theory of information~\cite{Shannon1948},
the amount of information contained in each state of a system with $N$
distinguishable states is $(\log N)$ bits of information. For example, the type
$\mathbb{2}$ can be thought of as an abstract system with two distinguishable
states labeled $\mathsf{true}$ and $\mathsf{false}$ each containing $\log 2 = 1$
bit. Similarly, the type $\mathbb{3}$ can be thought of as an abstract system
with three distinguishable states each containing $\log 3$ bits. The logarithmic
map implies that information contained in a composite state is the sum of the
information contained in its constituents. For example, the type $\mathbb{2}
\otimes \mathbb{3}$ can be thought of a composite system consisting of two
independent unrelated subsystems. Each state of the composite system therefore
contains $\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
information contained in each subsystem. Since all the combinators preserve
cardinality, they are also information-preserving.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^/$: Syntax and Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^/$: Additional Examples}

@@

@@

We have presented a natural notion of \emph{fractional types} that enriches a
class of reversible programming languages in several dimensions.  Further
research might show how to use fractional types in a conventional (i.e.,
irreversible) programming language, their full potential is only achieved when
the ambient language guarantees that no information is created or erased. The
key semantic insight is that iterating a reversible program $p$ on a finite type
must eventually reach the identity in $\ord{p}$ steps. By being careful not to
collapse proofs, each such reversible program has $\ord{p}$ distinct proofs of
reversibility and hence gives rise to a groupoid with cardinality
$\frac{1}{\ord{p}}$. Going from this observation to a full programming language
required several difficult and subtle design choices which we have explored to
produce $\Pi^/$. The latter language has a natural denotational semantics where
types denote groupoids. Its operational semantics requires a mechanism to
express a computational effect which enables spatially separated parts of the
program to communicate in a way that is reminiscent of entanglement in quantum
mechanics. It appears possible to realize such an operational semantics using
global reference cells, backtracking, or other conventional technique.

Our fractional types extend the natural denotation of types from sets to
non-trivial groupoids but they only scratch the surface of the tower of weak
$\omega$-groupoids that is expressible in HoTT. A long term goal of our research
is to find natural type constructors inspired by the rich combinatorial
structure of weak $\omega$-groupoids and that provide novel programming
abstractions. We conclude by outlining several interesting connections and
potential avenues for future work.

\paragraph*{Quotient Types.} Certain groupoids naturally correspond to
conventional \emph{quotient types}. Traditionally~\cite{quotient}, a quotient
type $\ag{T}{E}$ combines a type $T$ with an equivalence relation $E$ that
serves as the equality relation on the elements of $T$. Our notion of fractional
types in $\pifrac$ appears to subsume conventional quotient types and their
applications~\cite{Cohen2013} (e.g., defining fractions, multivariate
polynomials, field extensions, algebraic numbers, etc.) and it would interesting
to explicitly formalize this connection.

\paragraph*{Conservation of Information and Negative Entropy.} According to the
conventional theory of communication~\cite{Shannon1948}, a type with~$N$ values
is viewed as an abstract system that has~$N$ distinguishable states where the
amount of information contained in each state is $(\log{N})$. This entropy is a
measure of information which materializes itself in memory or bandwidth
requirements when storing or transmitting elements of this type. Thus a type
with 8 elements needs 3 bits of memory for storage or 3 bits of bandwidth for
communication. The logarithmic map implies that information contained in a
composite state is the sum of the information contained in its constituents. For
example, the type $\mathbb{2} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state of the
composite system therefore contains $\log (2 * 3) = \log 2 + \log 3$ bits which
is the sum of the information contained in each subsystem. If quantum field
theory is correct (as it so far seems to be) then information, during any
physical process, is neither created nor
destroyed. Landauer~\cite{Landauer:1961,bennett1985fundamental,Landauer},
Bennet~\cite{Bennett:1973:LRC,bennett2003notes,bennett2010notes},
Fredkin~\cite{fredkin1982conservative} and others made compelling arguments that
this physical principle induces a corresponding computational principle of
``conservation of information.'' In the context of finite types, generated from
the empty type $\zt$, the unit type $\ot$, and sums and products $\oplus$ and
$\otimes$, this principle states that the foundational (i.e., physical) notion
of computation is computation via type
isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
equivalences~\cite{Carette2016}, which are both sound and complete with respect
to cardinality-preserving maps. The introduction, in $\pifrac$, of types
(groupoids) with fractional cardinalities introduces types with \emph{negative
  entropy}. For example, a type with cardinality $\frac{1}{8}$ has entropy
$\log{\frac{1}{8}} = -3$. It is natural to interpret this negative entropy just
like we interpret ``negative money,'' as a debt to be repaid by some other part
of the system. In fact, the zig-zag example can be viewed as modeling a credit
card transaction where the money is produced at the output site by generating a
corresponding debt that is reconciled at the input site.

\paragraph*{Correspondence with Commutative Semifields.} Computations over
finite types naturally emerge from viewing types as syntax for semiring
elements, semiring identities as type isomorphisms, and justifications for
semiring identities as program transformations and
optimizations~\cite{Carette2016}. This correspondence provides a rich
proof-relevant version of the Curry-Howard correspondence between algebra and
reversible programming languages. The addition of full fractional types to the
mix would enrich the correspondence to commutative semifields, providing a
categorification~\cite{math/9802029} of the non-negative rational numbers in a
computational setting. This correspondence in $\Pi^/$ is still lacking,
however. Given any non-negative rational number, we can form a type whose
cardinality is that number. And yet, our types do not capture the full structure
of the non-negative rational numbers, as these form a commutative
semifield. There are a number of operations and properties which need to be
added to complete this.  Most seem quite straightforward.  We can however single
one out which may not be: what we are missing is a multiplicative inverse for
every type, and not just $\order{p}$. In particular, we would like to form the
type $\iorder{(\iorder{p})}$ and have it be equivalent to $\order{p}$.  As
explained in Sec.~\ref{sec:groupoids}, this would require a futher
generalization of looping and delooping that can be iterated. We leave a
possible extension with a general looping/delooping process to future work.
